The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2014-09-08 Papers/Universalis/ModalCollapse.tex
Improve text about modal collapse

\section{The Collapse of the Modal Square}

A crucial step of the ontological argument is the claim that if God's existence is possible, then it is necessary. This is Lemma L2 in G\"odel's proof. In the modal square of opposition (Fig. \ref{ToDo}), this is an unusual situation in which the \textbf{I} corner implies and entails the \textbf{A} corner, for the particular case when $\phi$ is $\exq x G(x)$. G\"odel's proof shows that his axioms are strong enough to invert the direction of entailment for the particular $\phi$ at issue. The question is whether the axioms are then not too strong, also allowing the inverted entailment for arbitrary $\phi$. That is essentially the question asked by Sobel \cite{Sobel1987}; and his proof of the modal collapse (MC) answers it affirmatively. The modal collapse is a collapse of the modal square: it shows that, assuming G\"odel's axioms, the subcontraries entail the contraries.

@@

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[MC] Everything that is the case is so necessarily:
  $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
\item[MC'] Everything that is possible is necessary:
  $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
\item[MC''] Modalities collapse:
  $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
\end{figure}

>>>>

\section{A Collapse of the Modal Square}

A crucial step of most ontological arguments is the claim that 
if God's existence is possible, then it is necessary. 
This is Lemma L2 in G\"odel's proof. 
In the modal square of opposition (Fig. \ref{ToDo}), 
this is an unusual situation in which the \textbf{I} corner 
must imply and entail the \textbf{A} corner, 
in the particular case when $\phi$ is $\exq x G(x)$. 
G\"odel's proof shows that his axioms are strong enough 
to invert the direction of entailment for the sentence at issue. 
The question, however, is whether the axioms are not too strong, 
also allowing the inverted entailment for arbitrary $\phi$. 
That is essentially the question asked by Sobel \cite{Sobel1987}; 
and his proof of the modal collapse (MC) provides an affirmative 
answer. It is possible to show that this form of the modal 
collapse entails (in modal logic K) a collapse of the modal 
square (MC'): the subcontraries entail (and even imply) the 
contraries. Furthermore, in any modal logic where where the 
axiom T holds (i.e. the accessibility relation is reflexive), 
a total collapse of the modalities (MC'') occurs.

@@

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[MC] Everything that is the case is so necessarily:
  $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
\item[MC'] Everything that is possible is necessary:
  $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
\item[T] Everything that is necessary is the case:
  $$\hol{\allq \phi [\nec \phi \imp \phi]}$$ 
\item[EI] (Modal Existential Import):
  $$\hol{\pos \top}$$ 
\item[AI] Everything that is necessary is possible:
  $$\hol{\allq \phi [\nec \phi \imp \pos \phi]}$$ 
\item[MC''] Modalities collapse completely:
  $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
\end{figure}
